Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-09-30 paper/current/main.tex
abstract and introduction

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of career trajectories capture the latent structure of global scientific mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by physical separation---geographical distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for an alternative modeling framework for human mobility that take these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
Here, using a massive dataset of scientific mobility and U.S. flight itineraries, we demonstrate a neural embedding model that successfully encodes nuanced relationships between locations into a vector-space ~\autocite{mikolov2013word2vec}, providing an effective functional distance that reflects the multi-faced structure of human mobility. 
Focusing on the case of scientific mobility between organizations, we show that the embedding captures cultural and linguistic relations, and even academic prestige at multiple levels of granularity, as well as implicit hierarchical structure.
Furthermore, vector representations of organizations themselves also reveal insights into the landscape of global scientific mobility, such as a relationship between organizations' size and the norm of its vector that is consistent across most countries. 
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly derived from the data can open up a new avenue of studying human mobility.

@@

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facets of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.  
Low dimensional distance alone cannot represent the multitude of inter-related factors that derive mobility. 
Networks offer a solution to representing the many dimensions of mobility, yet they only represent flows between discrete entities, and so cannot easily capture the relationships between entities themselves. 
Overcoming the complexity of mobility requires moving beyond simple functional distances and networks, to instead learning high-dimensional landscapes of mobility that incorporates the many facets of mobility into a single fine-grained and continuous representation.

Here, we propose using a neural embedding model that draws from real-world mobility trajectories can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationship based on culture, language, and even prestige along with geographic relationship can be probed. 
We leverage a massive dataset that documents scientific mobility captured in bibliometric records. Scientific mobility, which is central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on science of science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

>>>>

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of trajectories captures the latent structure of mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by physical separation---geographical distance~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for an alternative modeling framework for human mobility that takes these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
Here, using a massive dataset of scientific mobility and U.S. flight itineraries, we demonstrate that a neural embedding model can successfully encode nuanced relationships between locations into a vector-space ~\autocite{mikolov2013word2vec}, providing an effective functional distance that reflects the multi-faced structure of human mobility. 
Focusing on the case of scientific mobility, we show that the embedding captures cultural and linguistic relations at multiple levels of granularity, and even academic prestige.
Furthermore, embedding vectors themselves also reveal insights on the role of the organization in the landscape of global scientific mobility universally.
The ability to learn scalable, dense and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying mobility across domains.

@@

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.  
Low dimensional distance alone cannot represent the multitude of inter-related factors that derive mobility. 
Networks offer a solution to representing many dimensions of mobility, yet they only represent flows between discrete entities, and so cannot easily capture the relationships between entities themselves. 
Overcoming the complexity of mobility requires moving beyond simple functional distances and networks, to instead learning high-dimensional landscapes of mobility that incorporates the many facets of mobility into a single fine-grained and continuous representation.

Here, we propose to apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationships based on culture, language, and even prestige along with geographic relationship can be probed. 
We leverage two massive datasets: a dataset of scientific mobility captured in bibliometric records and a dataset of US airport itinerary records. In particular, we focus on scientific mobility throughout the paper due to its richness and importance. Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on the science of science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.
